---
title: "Report"
output: html_document
date: "2023-03-11"
---


We want to 
We are scanning sequences to distinguish patters(motifs) in a dataset


As our foreground, we will be using a DNA subsequence on which a protein may bind(Transcription Factor Binding Site)
We are searching for motifs, in a region of DNA, which is responsible for regulating the (expression level of a gene)








We have the gene names of the Glycolosis and Glyconeogenesis pathways.
We want to investigate if these 2 lists have more common genes than expected by chance.

```{r}

#Read data 
glycoly_gene_ids = read.table("gene_id_pathway//Glycolysis_path.txt", header=TRUE, sep ="\n")
glyconeo_gene_ids = read.table("gene_id_pathway//Gluconeogenesis_path.txt", header=TRUE, sep="\n")

#find intersection of lists
common_genes = intersect(unlist(glycoly_gene_ids[1]),unlist(glyconeo_gene_ids[1]))

print(common_genes)


```

We now want to import our foreground for each pathway.


```{r}

getData = function(filepath) {
  con = file(filepath, "r")
  data = list()
  while ( TRUE ) {
    line = readLines(con, n = 1)
    if ( length(line) == 0 ) {
    break
    }
    if(length(grep(">", line)) > 0){
    name = gsub("^>(\\w+)", replacement="\\1", x=line)
    data[[name]] = ""
  }
  else{
  data[[name]] = paste(data[[name]], line, sep="")
  }
}
close(con)
return(data)
}




glycoly_foreground = getData("foreground//Glycolysis_foreground.txt")
glyconeo_foreground = getData("foreground//Glycolysis_foreground.txt")
```









